The sound of teaching music: Expert pianists' performance modulations for novices

### 2022-08-20 paper/manuscript.Rmd
update references and after intro
Signed-off-by: Atsuko Tominaga <Tominaga_Atsuko@phd.ceu.edu>

If experts rely on generic action modulations, it can be expected that they will play more slowly during didactic demonstration, regardless of the kind of techniques to be taught. To the extent that experts use action modulations to support their teaching of specific techniques, they should exaggerate legato (i.e., key-overlap time should be more positive) and staccato (i.e., key-overlap time should be more negative) when teaching articulation whereas they should exaggerate forte (i.e., key velocity should be higher) and piano (i.e., key velocity should be lower) when teaching dynamics. Furthermore, one could speculate that they might produce modulations specifically at structurally important points that best highlight the technique to be taught. Importantly, they should avoid modulating irrelevant properties of expression, e.g., the smoothness of sound while teaching dynamics.

@@

In the teaching condition, participants were instructed to play the excerpt of music as if they were teaching it to students. It was mentioned that the students already knew the sequence of the tones and that they were trying to learn how to perform the piece with the notated expression by listening to the participant's performance. In the performing condition, participants were asked to play the excerpt of music as if they were performing it to an audience (see details in *[Supplementary material](#supplementary)*). Participants played the piece 8 times per expressive technique per condition, so there were 32 trials in total (2 conditions x 2 expressive techniques x 8 trials). The order of the conditions was blocked and counterbalanced across participants. The order of the expressive techniques within each condition was also blocked and counterbalanced across participants. A leading metronome (80 quarter beats per minute, 8 beats) indicated the target tempo before each trial.

@@

Data cleaning, preprocessing and statistical analysis were performed in R version 4.0.5. For statistical analysis, only 16th notes with expressive notations were included. Overall, five trials were excluded from data analysis because participants did not follow the sheet music or stopped performing before the end. Pitch errors were identified by comparing the sequence of musical notes produced by a participant with the sequence of musical notes according to the sheet music. Pitch errors included either extra, missing or substituted tones and were manually removed by using the *editData* R package. For onsets, `r round(sum(error_onset_summary$N)/((31*4*8)-5)*100, 2)`% of the trials contained at least one pitch error (extra notes: `r round(error_onset_summary[Reason == "Extra Notes"]$N/((31*4*8)-5)*100, 2)`%, missing notes: `r round(error_onset_summary[Reason == "Missing Notes"]$N/((31*4*8)-5)*100, 2)`%, substituted notes: `r round(error_onset_summary[Reason == "Substituted"]$N/((31*4*8)-5)*100, 2)`%). For offsets, `r round(sum(error_offset_summary$N)/((31*4*8)-5)*100, 2)`% of the trials contained at least one pitch error (extra notes: `r round(error_offset_summary[Reason == "Extra Notes"]$N/((31*4*8)-5)*100, 2)`%, missing notes: `r round(error_offset_summary[Reason == "Missing Notes"]$N/((31*4*8)-5)*100, 2)`%, substituted notes: `r round(error_offset_summary[Reason == "Substituted"]$N/((31*4*8)-5)*100, 2)`%). We found that some participants did not follow sheet music (e.g., hold some notes longer than notated), therefore the order of offsets did not correspond to that of onsets. We counted these as errors and removed the erroneous notes even if the order of onsets was correct. As a result, less than 1 % of total responses were corrected. In addition to pitch errors, we removed outliers for IOIs, KOT and KV, defined as values more than 3 standard deviations from the mean of each dependent variable. For each dependent variable, this resulted in less than 5% of overall responses being removed as outliers.

We performed separate analyses for the two expressive techniques (i.e., articulation and dynamics). A paired-sample *t*-test or a Wilcoxon Signed-rank test (a non-parametric alternative to a paired *t*-test) was performed to compare the mean IOIs in the teaching and performing condition. For KOT and KV, we performed a 2 x 2 repeated-measures analysis of variance (ANOVA) with the factors Condition (teaching vs. performing) and Subcomponent (Articulation: legato vs. staccato or Dynamics: forte vs. piano, respectively). The *t.test* or *wilcox.test* function in the *stats* R package was used for a *t*-test or a Wilcoxon Signed-rank test. For calculating an effect size, we used the *rstatix* R package. The *aov_car* function in the *afex* R package was used for a repeated-measures ANOVA. For post-hoc comparisons on the estimated marginal means, we used the *emmeans* R package.

@@

All effects are reported as significant at *p* < .05. We first report results for performance of the piece with the notated expression of articulation (*Fig \ref{fig:stimuli} B*), followed by performance of the piece with the notated expression of dynamics (*Fig \ref{fig:stimuli} C*). To recall our predictions, if participants play more slowly when they are trying to teach, inter-onset intervals (IOIs) should be larger when teaching. If participants specifically modulate relevant aspects of the expressive techniques they are trying to teach, key-overlap time (KOT) should be more positive for legato and more negative for staccato when teaching articulation, and key velocity (KV) should be higher for forte and lower for piano when teaching dynamics.

@@

The results for KOT and KV showed that participants successfully highlighted relevant aspects of articulation and dynamics. Specifically, participants exaggerated staccato when teaching articulation and forte when teaching dynamics. Moreover, our exploratory analysis demonstrated that participants made a larger contrast in dynamics at transition points (i.e., forte to piano or piano to forte). Importantly, participants did not modulate their performance in terms of irrelevant aspects of the techniques for teaching purposes (e.g., modulating the smoothness of sound while teaching dynamics). These findings confirmed that participants modulated their performance in systematic and fine-grained ways to teach expressive techniques.

@@

The same apparatus as in Experiment 1 was used. We selected Clementi's Sonatina Op.36 (No.3) in C major as a stimulus because it contained our targeted expressions (i.e., articulation and dynamics) and was relatively simple in terms of motor skills. The first 12 measures of the original piece were used and modified so that the piece had an almost equal number of data points for each dependent variable. The modified piece consisted of a 12-measure isochronous melody notated in 4/4 meter to be played with the right hand only. Original sheet music was used for the purpose of practice (*Fig \ref{fig:stimuli-2} A*). Expressive notations were added to the original sheet music for the experiment (*Fig \ref{fig:stimuli-2} B-C*). These excerpts were confirmed to be musically natural by a doctoral student in piano performance at Liszt Ferenc Academy of Music in Hungary. The fingering was also assigned and confirmed by the same doctoral student.

@@

We employed the same procedure as in Experiment 1 with several modifications. First, all participants were required to memorise the piece without expressive notations (i.e., *Fig \ref{fig:stimuli-2} A*) prior to the experiment so that they had enough time to familiarise themselves with the piece and perform it without pitch errors while performing it with different notated expressions in the lab. Second, we modified the wording of instructions for the performing condition so that both instructions had the same focus on expressive notations because there was less emphasis on expression in the instruction in the performing condition in Experiment 1 (see details in *[Supplementary material](#supplementary)*). Third, participants could choose their preferred tempo from one of three options (100, 110 and 120 quarter beats per minute) because some evidence show that musicians have their own preferred tempo, which affects their temporal variability [@zamm_2018]. The chosen tempo was again cued by a leading metronome.  To make sure that participants memorised the piece and had sufficient motor skills, we asked participants to perform the piece without looking at the original sheet music (*Fig \ref{fig:stimuli-2} A*). Those who could not perform the piece without pitch errors twice consecutively within five attempts were excluded from data analysis (as a result, one participant was excluded). The rest of the procedure was identical to Experiment 1.

@@

Data cleaning, preprocessing and statistical analysis were almost identical to Experiment 1. For statistical analysis, only 8th notes with expressive notations were included. As a result, only one 8th note in the 4th measure without any expression was not included. IOIs were normalised by their preferred tempo because each participant chose a tempo from the three options. Given the different tempi, key-overlap ratios (KORs) were calculated by dividing KOT by the mean IOI of each performance to normalise KOT. Additionally, we include KV difference (i.e., KV difference for each interval) at transition points (i.e., forte to piano or piano to forte) as a dependent variable based on the findings of Experiment 1. Three trials were entirely excluded from data analysis because participants did not follow the sheet music. Using the same approach as in Experiment 1, pitch errors were removed manually. For onsets, `r round(sum(error_onset_summary_2$N)/((20*4*8)-3)*100, 2)`% of the trials contained at least one pitch error (extra notes: `r round(error_onset_summary_2[Reason == "Extra Notes"]$N/((20*4*8)-3)*100, 2)`%, missing notes: `r round(error_onset_summary_2[Reason == "Missing Notes"]$N/((20*4*8)-3)*100, 2)`%, substituted notes: `r round(error_onset_summary_2[Reason == "Substituted"]$N/((20*4*8)-3)*100, 2)`%). For offsets, `r round(sum(error_offset_summary_2$N)/((20*4*8)-3)*100, 2)`% of the trials contained at least one pitch error (extra notes: `r round(error_offset_summary_2[Reason == "Extra Notes"]$N/((20*4*8)-3)*100, 2)`%, missing notes: `r round(error_offset_summary_2[Reason == "Missing Notes"]$N/((20*4*8)-3)*100, 2)`%, substituted notes: `r round(error_offset_summary_2[Reason == "Substituted"]$N/((20*4*8)-3)*100, 2)`%). As a result, less than 1 % of total responses were corrected. For each dependent variable, removing outliers (i.e., responses outside 3 standard deviations from the mean) resulted in less than 5% of overall responses being removed.

@@

The results from Experiment 2 replicate our earlier findings and provide further evidence that skilled pianists were able to modulate their performance for teaching purposes. As in Experiment 1, we observed slower performance only while teaching the notated articulation. Again, KORs showed that when teaching articulation, participants exaggerated staccato while there was no significant difference in legato between the two conditions. While teaching dynamics, KV showed that participants exaggerated forte and piano. This pattern of exaggeration was more pronounced than in Experiment 1, where only higher KV for forte was observed. Furthermore, we again found that when teaching dynamics, at transition points participants produced a larger contrast in dynamics between forte and piano, bidirectionally. Importantly, participants did not modulate performance aspects that were irrelevant for the techniques to be taught (i.e., no modulation of dynamics when teaching articulation and vice versa). Overall, Experiment 2 demonstrated systematic and fine-grained didactic performance modulations for a naturalistic piece of music.

@@

The aim of the study was to investigate how expert pianists modulate their performance to teach expressive techniques. Overall, the findings of the two experiments showed that pianists performed one and the same piece differently depending on whether they played with the intention to teach expressive techniques or with the intention to perform the piece for an audience. When playing with the intention to teach, pianists selectively highlighted relevant aspects of artistic expressions. Across two different pieces, participants exaggerated staccato when playing with the intention to teach expressions concerning articulation. The lack of a difference for legato when teaching articulation might stem from a ceiling effect. When playing with the intention to teach expressions concerning dynamics, participants made larger dynamics changes between forte and piano (in both directions). This constitutes an effective way to highlight the technique as loudness is determined relatively. Taken together, our findings demonstrate that expert pianists systematically modulated their sound depending on the specific technique they were trying to convey.

Although participants tended to play slower in the teaching condition than in the performing condition in general, we found a significant difference only for articulation, not for dynamics. As mentioned in the discussion of Experiment 1, it could be that the metronome beats given prior to each performance discouraged participants from deviating from the prescribed tempo. Another possibility is that expression is best taught when leaving the tempo unchanged when it is irrelevant. This would imply that general didactic modulations like slowing down might be less useful in the context of teaching expressive techniques where timing itself can be used to add an expression to a performance.

Musicians and music teachers tend to consider expressive skills as a performer’s most important skill [@lindstrom_2003]. In the present study, experts focused on the teaching of expressions that were notated. However, expression also has other facets [@juslin_2003] that are not only piece-related but performer- or context-related. An interesting topic for future research is how experts teach not only expressive techniques as such but convey possibilities of interpretation.

One limitation of the study is that although we instructed participants to imagine a situation in which they were teaching musical expression to students, there was no feedback from actual students. Given that teachers modulate their demonstration throughout online interactions with learners [@okazaki_2019], future studies are needed to investigate how experts dynamically adapt their performance to their students' skill level and demonstrated abilities.

The present findings extend earlier research on teaching-related action modulations. First, our study sheds light on the teaching of expressive techniques that are an integral part of skill acquisition in artistic domains. We showed that compared to an expressive performance baseline, experts made specific modulations to teach particular techniques. Second, the specificity of the observed modulations supports the idea that teaching comprises more than generic modulations like slowing down or overall exaggeration that may draw learners' attention. Rather, expert demonstrators seem to follow principles of relevance in communication [@sperber_1995], highlighting only those aspects they are intending to demonstrate. How learners benefit from the perceptual and motor cues that come with specific exaggerations, and whether understanding the teacher's intentions explicitly adds to the learning success are important questions for future research.

>>>>

If experts rely on generic action modulations, it can be expected that they will play more slowly during didactic demonstration, regardless of the kind of techniques to be taught. To the extent that experts use action modulations to support their teaching of specific techniques, they should exaggerate legato (i.e., key-overlap time should be more positive) and staccato (i.e., key-overlap time should be more negative) when teaching articulation whereas they should exaggerate forte (i.e., key velocity should be higher) and piano (i.e., key velocity should be lower) when teaching dynamics. Furthermore, one could speculate that they might produce modulations specifically at structurally important points that best highlight the technique to be taught. Importantly, they should avoid modulating irrelevant properties of expressio (e.g., the smoothness of sound while teaching dynamics).

@@

In the teaching condition, participants were instructed to play the excerpt of music as if they were teaching it to students. It was mentioned that the students already knew the sequence of the tones and that they were trying to learn how to perform the piece with the notated expression by listening to the participant's performance. In the performing condition, participants were asked to play the excerpt of music as if they were performing it to an audience (see details in *[Supplementary material](#supplementary)*). Participants played the piece 8 times per technique per condition, so there were 32 trials in total (2 conditions x 2 techniques x 8 trials). The order of the conditions was blocked and counterbalanced across participants. The order of the techniques within each condition was also blocked and counterbalanced across participants. A leading metronome (80 quarter beats per minute, 8 beats) indicated the target tempo before each trial.

@@

Data cleaning, preprocessing and statistical analysis were performed in R version 4.0.5. For statistical analysis, only 16th notes with expressive notations were included. Overall, five trials were excluded from data analysis because participants did not follow the sheet music or stopped performing before the end. Pitch errors were identified by comparing the sequence of musical notes produced by a participant with the sequence of musical notes according to the sheet music. Pitch errors included either extra, missing or substituted tones and were manually removed by using the *editData* R package. For onsets, `r round(sum(error_onset_summary$N)/((31*4*8)-5)*100, 2)`% of the trials contained at least one pitch error (extra notes: `r round(error_onset_summary[Reason == "Extra Notes"]$N/((31*4*8)-5)*100, 2)`%, missing notes: `r round(error_onset_summary[Reason == "Missing Notes"]$N/((31*4*8)-5)*100, 2)`%, substituted notes: `r round(error_onset_summary[Reason == "Substituted"]$N/((31*4*8)-5)*100, 2)`%). For offsets, `r round(sum(error_offset_summary$N)/((31*4*8)-5)*100, 2)`% of the trials contained at least one pitch error (extra notes: `r round(error_offset_summary[Reason == "Extra Notes"]$N/((31*4*8)-5)*100, 2)`%, missing notes: `r round(error_offset_summary[Reason == "Missing Notes"]$N/((31*4*8)-5)*100, 2)`%, substituted notes: `r round(error_offset_summary[Reason == "Substituted"]$N/((31*4*8)-5)*100, 2)`%). We found that some participants did not precisely follow the sheet music (e.g., they held some notes longer than notated), therefore the order of offsets did not correspond to that of onsets. We counted these as errors and removed the erroneous notes even if the order of onsets was correct. As a result, less than 1 % of total responses were corrected. In addition to pitch errors, we removed outliers for IOIs, KOT and KV, defined as values more than 3 standard deviations from the mean of each dependent variable. For each dependent variable, this resulted in less than 5% of overall responses being removed as outliers.

We performed separate analyses for the two techniques (i.e., articulation and dynamics). A paired-sample *t*-test or a Wilcoxon Signed-rank test (a non-parametric alternative to a paired *t*-test) was performed to compare the mean IOIs in the teaching and performing condition. For KOT and KV, we performed a 2 x 2 repeated-measures analysis of variance (ANOVA) with the factors Condition (teaching vs. performing) and Subcomponent (Articulation: legato vs. staccato or Dynamics: forte vs. piano, respectively). The *t.test* or *wilcox.test* function in the *stats* R package was used for a *t*-test or a Wilcoxon Signed-rank test. For calculating an effect size, we used the *rstatix* R package. The *aov_car* function in the *afex* R package was used for a repeated-measures ANOVA. For post-hoc comparisons on the estimated marginal means, we used the *emmeans* R package.

@@

All effects are reported as significant at *p* < .05. We first report results for performance of the piece with the notated expression of articulation (*Fig \ref{fig:stimuli} B*), followed by performance of the piece with the notated expression of dynamics (*Fig \ref{fig:stimuli} C*). To recall our predictions, if participants play more slowly when they are trying to teach, inter-onset intervals (IOIs) should be larger when teaching. If participants specifically modulate relevant aspects of the techniques they are trying to teach, key-overlap time (KOT) should be more positive for legato and more negative for staccato when teaching articulation, and key velocity (KV) should be higher for forte and lower for piano when teaching dynamics.

@@

The results for KOT and KV showed that participants successfully highlighted relevant aspects of articulation and dynamics. Specifically, participants exaggerated staccato when teaching articulation and forte when teaching dynamics. Moreover, our exploratory analysis demonstrated that participants made a larger contrast in dynamics at transition points (i.e., forte to piano or piano to forte). Importantly, participants did not modulate their performance in terms of irrelevant aspects of the techniques for teaching purposes (e.g., modulating the smoothness of sound while teaching dynamics). These findings confirmed that participants modulated their performance in systematic and fine-grained ways to teach particular techniques.

@@

The same apparatus as in Experiment 1 was used. We selected Clementi's Sonatina Op.36 (No.3) in C major as a stimulus because it contains our targeted expressions (i.e., articulation and dynamics) and is relatively simple in terms of motor skills. The first 12 measures of the original piece were used and modified so that the piece had an almost equal number of data points for each dependent variable. The modified piece consisted of a 12-measure isochronous melody notated in 4/4 meter to be played with the right hand only. Original sheet music was used for the purpose of practice (*Fig \ref{fig:stimuli-2} A*). Expressive notations were added to the original sheet music for the experiment (*Fig \ref{fig:stimuli-2} B-C*). These excerpts were confirmed to be musically natural by a doctoral student in piano performance at Liszt Ferenc Academy of Music in Hungary. The fingering was also assigned and confirmed by the same doctoral student.

@@

We employed the same procedure as in Experiment 1 with several modifications. First, all participants were required to memorise the piece without expressive notations (i.e., *Fig \ref{fig:stimuli-2} A*) prior to the experiment so that they had enough time to familiarise themselves with the piece and perform it without pitch errors while performing it with different notated expressions in the lab. Second, we modified the wording of instructions for the performing condition so that both instructions had the same focus on expressive notations because there had been less emphasis on expression in the instruction in the performing condition in Experiment 1 (see details in *[Supplementary material](#supplementary)*). Third, participants could choose their preferred tempo from one of three options (100, 110 and 120 quarter beats per minute) because some evidence show that musicians have their own preferred tempo, which affects their temporal variability [@zamm_2018]. The chosen tempo was again cued by a leading metronome.  To make sure that participants memorised the piece and had sufficient motor skills, we asked participants to perform the piece without looking at the original sheet music (*Fig \ref{fig:stimuli-2} A*). Those who could not perform the piece without pitch errors twice consecutively within five attempts were excluded from data analysis (as a result, one participant was excluded). The rest of the procedure was identical to Experiment 1.

@@

Data cleaning, preprocessing and statistical analysis were almost identical to Experiment 1. For statistical analysis, only 8th notes with expressive notations were included. As a result, only one 8th note in the 4th measure without any expression was not included. IOIs were normalised by their preferred tempo because each participant chose a tempo from the three options. Given the different tempi, key-overlap ratios (KORs) were calculated by dividing KOT by the mean IOI of each performance to normalise KOT. Additionally, we included KV difference (i.e., KV difference for each interval) at transition points (i.e., forte to piano or piano to forte) as a dependent variable based on the findings of Experiment 1. Three trials were entirely excluded from data analysis because participants did not follow the sheet music closely enough. Using the same approach as in Experiment 1, pitch errors were removed manually. For onsets, `r round(sum(error_onset_summary_2$N)/((20*4*8)-3)*100, 2)`% of the trials contained at least one pitch error (extra notes: `r round(error_onset_summary_2[Reason == "Extra Notes"]$N/((20*4*8)-3)*100, 2)`%, missing notes: `r round(error_onset_summary_2[Reason == "Missing Notes"]$N/((20*4*8)-3)*100, 2)`%, substituted notes: `r round(error_onset_summary_2[Reason == "Substituted"]$N/((20*4*8)-3)*100, 2)`%). For offsets, `r round(sum(error_offset_summary_2$N)/((20*4*8)-3)*100, 2)`% of the trials contained at least one pitch error (extra notes: `r round(error_offset_summary_2[Reason == "Extra Notes"]$N/((20*4*8)-3)*100, 2)`%, missing notes: `r round(error_offset_summary_2[Reason == "Missing Notes"]$N/((20*4*8)-3)*100, 2)`%, substituted notes: `r round(error_offset_summary_2[Reason == "Substituted"]$N/((20*4*8)-3)*100, 2)`%). As a result, less than 1 % of total responses were corrected. For each dependent variable, removing outliers (i.e., responses outside 3 standard deviations from the mean) resulted in less than 5% of overall responses being removed.

@@

The results from Experiment 2 replicate our earlier findings and provide further evidence that skilled pianists modulated their performance in specific ways for teaching purposes. As in Experiment 1, we observed slower performance only while teaching the notated articulation. Again, KORs showed that when teaching articulation, participants exaggerated staccato while there was no significant difference in legato between the two conditions. While teaching dynamics, KV showed that participants exaggerated forte and piano. This pattern of exaggeration was more pronounced than in Experiment 1, where only higher KV for forte was observed. Furthermore, we again found that when teaching dynamics, at transition points participants produced a larger contrast in dynamics between forte and piano, bidirectionally. Importantly, participants did not modulate performance aspects that were irrelevant for the techniques to be taught (i.e., no modulation of dynamics when teaching articulation and vice versa). Overall, Experiment 2 demonstrated systematic and fine-grained didactic performance modulations for a naturalistic piece of music.

@@

The aim of this study was to investigate how expert pianists modulate their performance to teach expressive techniques. Overall, the findings of the two experiments showed that pianists performed one and the same piece differently depending on whether they played with the intention to teach expressive techniques or with the intention to perform the piece for an audience. When playing with the intention to teach, pianists selectively highlighted relevant aspects of artistic expressions. Across two different pieces, participants exaggerated staccato when playing with the intention to teach expressions concerning articulation. The lack of a difference for legato when teaching articulation might stem from a ceiling effect because there might not be enough room for exaggeration without significantly dropping the tempo. When playing with the intention to teach expressions concerning dynamics, participants made larger dynamics changes between forte and piano (in both directions). This constitutes an effective way to highlight the technique as loudness is determined relatively. Taken together, our findings demonstrate that expert pianists systematically modulated their sound depending on the specific technique they were trying to convey.

Although participants tended to play slower in the teaching condition than in the performing condition in general, we found a significant difference only for articulation, not for dynamics. As mentioned in the discussion of Experiment 1, it could be that the metronome beats given prior to each performance discouraged participants from deviating from the prescribed tempo. Another possibility is that expression is best taught when leaving the tempo unchanged when it is irrelevant. This would imply that general didactic modulations like slowing down might be less useful in the context of teaching techniques where timing itself can be used to add an expression to a performance.

One limitation of the study is that although we instructed participants to imagine a situation in which they were teaching musical expression to students, there was no feedback from actual students. It can be expected that teachers would also make didactic modulations in some domains other than acoustic properties, such as gestures if they are physically present in front of students. Moreover, given that teachers modulate their demonstration throughout ongoing interactions with learners [@okazaki_2019], future studies are needed to investigate how experts dynamically adapt their performance to their students’ skill level and demonstrated abilities. Also, some of the participants in the current study did not have teaching experience at all and even those who had teaching experience had taught for only a few years. It would be important to investigate if our findings would be observed in participants with extensive experience teaching the piano.

Another limitation of the study is that it is not certain how the results might generalise to performances of more complex pieces and pieces with fewer notations. In more complex pieces of music, multiple expressive notations could be assigned to one note or phrase. Also, it is rare that almost all the notes are assigned to one expressive notation as in our stimuli. Future studies are needed to determine whether pianists would exaggerate specific parts of a piece or prioritise one aspect over the others given multiple concurrent expressive notations and more possibilities for giving the music different interpretations.

Musicians and music teachers tend to consider expressive skills as performers’ most important skills [@lindstrom_2003]. In the present study, experts focused on the teaching of expressions that were notated. However, expression also has other facets [@juslin_2003] that are not only piece-related (e.g., notated expressions) but performer- or context-related (e.g., performer’s interpretation, listening contexts such as recordings or concerts). The ultimate goal of expressive performance is considered to lie in developing performers’ own interpretations of music and convey affects and emotions by using expressive tools such as articulation, dynamics, tempo and timing [@meissner_2021]. The current research opens up the possibility of studying the teaching of expressive performance by demonstrating that pianists could signal didactic intentions without relying on verbal instructions, which are heavily used in actually teaching settings [e.g., @woody_2000]. It would be interesting to examine how teachers play (i.e., demonstrate) relates to verbal instructions they give.

The present findings extend earlier research on teaching-related action modulations. First, our study sheds light on the teaching of technical skills that are an integral part of skill acquisition in artistic domains. We showed that compared to an expressive performance baseline, experts made specific modulations to teach particular techniques. Second, the specificity of the observed modulations supports the idea that teaching comprises more than generic modulations like slowing down or overall exaggeration that may draw learners’ attention. Rather, expert demonstrators seem to follow principles of relevance in communication [@sperber_1995], highlighting only those aspects they are intending to demonstrate. How learners benefit from the perceptual and motor cues that come with specific exaggerations, and whether understanding the teacher's intentions explicitly adds to the learning success are important questions for future research.
